% =============================================================================
% CENG 467 – Natural Language Understanding and Generation
% Take-Home Midterm Examination
% Student ID: 310201050
% Instructor: Prof. Dr. Aytuğ Onan
% Date: April 2026
% =============================================================================

\documentclass[11pt,a4paper]{article}

% ── Layout & encoding ────────────────────────────────────────────────────────
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[a4paper, margin=2.5cm]{geometry}
\usepackage{microtype}
\usepackage{setspace}
\onehalfspacing

% ── Mathematics ──────────────────────────────────────────────────────────────
\usepackage{amsmath, amssymb, amsthm}

% ── Tables / figures ─────────────────────────────────────────────────────────
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{array}

% ── Code listings (optional) ─────────────────────────────────────────────────
\usepackage{listings}
\usepackage{xcolor}
\definecolor{codegray}{gray}{0.95}
\lstset{
    basicstyle      = \ttfamily\small,
    backgroundcolor = \color{codegray},
    breaklines      = true,
    frame           = single,
    numbers         = left,
    numberstyle     = \tiny,
    keywordstyle    = \color{blue},
    commentstyle    = \color{gray},
    stringstyle     = \color{red!70!black},
    showstringspaces= false,
}

% ── Hyperlinks ───────────────────────────────────────────────────────────────
\usepackage[colorlinks=true, linkcolor=black, citecolor=blue, urlcolor=blue]{hyperref}
\usepackage{url}

% ── Bibliography ─────────────────────────────────────────────────────────────
\usepackage[numbers, sort&compress]{natbib}

% =============================================================================
% TITLE PAGE
% =============================================================================
\title{
    \vspace{-1cm}
    \large CENG 467 — Natural Language Understanding and Generation \\
    \vspace{0.4cm}
    \LARGE\textbf{Take-Home Midterm Examination} \\
    \vspace{0.6cm}
    \large A Comparative Study of Representation, Sequence Labeling, \\
    Summarization, Translation and Language Modeling
}
\author{
    \textbf{Student ID:} 310201050 \\
    \vspace{0.2cm}
    \small Instructor: Prof.\ Dr.\ Aytu\u{g} Onan \\
    \small \.{I}zmir Institute of Technology --- Department of Computer Engineering
}
\date{April 2026}

% =============================================================================
\begin{document}
\maketitle

\thispagestyle{empty}
\vspace{0.5cm}

\begin{abstract}
\noindent
This report presents a controlled comparative study of five core
NLP tasks: text classification, named entity recognition,
text summarization, machine translation, and language modeling.
For each task, two or three model families are evaluated on a
shared benchmark dataset under fixed random seeds, consistent
preprocessing pipelines, and held-out test sets that are used only
once for final evaluation. Sparse, dense, and contextual
representations are compared head-to-head, and qualitative error
analyses are provided alongside the quantitative results. Across
all five tasks, contextual representations from pretrained
Transformer models outperform classical and from-scratch recurrent
baselines, with the F1 gap reaching nearly 20 points for named
entity recognition and a 10$\times$ perplexity reduction for
language modelling. All experiments are reproducible from the
accompanying code repository on an Apple M2 (8\,GB) device with
seed 3102.
\end{abstract}

\vspace{0.5cm}
\noindent
\textbf{Repository:}
\url{https://github.com/<your-github-username>/CENG467_Midterm_310201050}

\vspace{1cm}
\tableofcontents

\newpage

% =============================================================================
% INTRODUCTION
% =============================================================================
\section{Introduction}

This document presents the solutions to the five questions of the
CENG 467 take-home midterm. Each section is self-contained, including
its own dataset description, methodology, hyperparameters, results,
and discussion.

The study is conducted under three overarching constraints. First,
all experiments use a fixed random seed (3102, derived from the
author's student identifier) to ensure exact reproducibility.
Second, dataset splits are kept consistent between models within a
single question; the test set is touched exactly once at the end of
each pipeline. Third, all training and inference is performed on a
single Apple~M2 device with 8\,GB of RAM, which informs the choice
of distilled and small-footprint models throughout the study.

The complete source code is organised by question
(\texttt{q1\_classification/}, \texttt{q2\_ner/}, etc.) and is
available in the accompanying GitHub repository. Reproduction
instructions are provided in the \texttt{README.md} file.

\subsection{Headline Results}

Table~\ref{tab:headline} summarises the principal finding from each
question. In every case, the contextual or pretrained Transformer
model outperforms the from-scratch baseline; the gap is largest where
training data is most limited (NER) and on tasks where surface
features are weakest (language modelling).

\begin{table}[h]
\centering
\caption{Best result on each task and the from-scratch baseline.}
\label{tab:headline}
\small
\begin{tabular}{llcc}
\toprule
Task & Headline metric & Baseline & Best (pretrained) \\
\midrule
Q1 Classification (IMDb)        & Macro-F1   & 0.7488 (BiLSTM)       & \textbf{0.8970} (DistilBERT) \\
Q2 NER (CoNLL-2003)             & F1         & 0.7011 (BiLSTM-CRF)   & \textbf{0.8975} (DistilBERT) \\
Q3 Summarization (CNN/DM)       & ROUGE-L    & 0.1883 (LexRank)      & \textbf{0.2580} (DistilBART) \\
Q4 Translation (Multi30k de–en) & BLEU       & 27.69 (Seq2Seq)       & \textbf{39.66} (MarianMT)    \\
Q5 Language Modeling (WikiText-2) & Test PPL & 3587.97 (Trigram)     & \textbf{353.41} (LSTM)       \\
\bottomrule
\end{tabular}
\end{table}

% =============================================================================
% INCLUDE EACH QUESTION
% =============================================================================
\input{q1_section}
\newpage
\input{q2_section}
\newpage
\input{q3_section}
\newpage
\input{q4_section}
\newpage
\input{q5_section}

% =============================================================================
% CONCLUSION
% =============================================================================
\newpage
\section{Conclusion}

Across the five tasks studied, a consistent pattern emerges:
contextual sub-word representations from pretrained Transformers
outperform both classical statistical baselines and recurrent
architectures trained from scratch, often by a substantial margin.
The advantage is most pronounced when training data is limited
(NER, MT) and when surface lexical features are weakly informative
(language modelling). On the IMDb classification task, where the
6{,}000-example training subset is small but the unigram signal is
strong, the sparse TF-IDF baseline came surprisingly close to
DistilBERT, even outperforming the from-scratch BiLSTM.

The study also highlights three trade-offs that should temper any
universal endorsement of contextual representations. First,
inference cost: pretrained Transformers are dramatically more
expensive than $n$-gram or LSTM baselines, a fact that became
concrete during these experiments when the DistilBERT fine-tune for
Q1 alone consumed roughly 45\,minutes on an M2. Second,
faithfulness: abstractive Transformer decoders can hallucinate
content not present in the source, a failure mode that extractive
baselines structurally cannot exhibit (Q3, example 50). Third,
transparency: a from-scratch model whose parameters one has
trained, debugged, and inspected is qualitatively different from a
downloaded checkpoint, and the educational and diagnostic value of
the former is non-trivial. The Q4 Seq2Seq model exhibited a
classic ``repetition trap'' (\textit{``a a a a a''}) in three of
three qualitative samples, a failure mode that is invisible when
one only consumes a pretrained black box.

Reproducibility was treated as a first-class constraint throughout.
Every result in this report can be regenerated by running the
provided scripts on a comparable device with seed 3102. The
associated repository contains the JSON output files used to
populate every numeric cell in this report.

% =============================================================================
% REFERENCES
% =============================================================================
\newpage
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}