% ─────────────────────────────────────────────────────────────────
% CENG 467 Midterm – Question 4: Machine Translation
% Student ID: 310201050
% ─────────────────────────────────────────────────────────────────

\section{Question 4: Machine Translation}

\subsection{Dataset and Preprocessing}

The Multi30k corpus \cite{elliott2016multi30k} is used for
German-to-English translation. Multi30k is built on top of the
Flickr30k image-caption dataset, with each English caption
professionally translated into German. It contains 29{,}000 training,
1{,}014 validation, and 1{,}000 test parallel sentence pairs. The
relatively short, syntactically-controlled captions make the corpus a
popular benchmark for neural-MT experiments on modest hardware.

Two distinct preprocessing pipelines are used:

\begin{itemize}
  \item \textbf{Seq2Seq pipeline:} text is lowercased, punctuation
        tokens are separated from adjacent words via regular
        expressions, and resulting tokens are mapped to integer ids
        using vocabularies built from the training data. Vocabulary
        size is capped at 8{,}000 per language with a minimum-frequency
        threshold of 2; rare tokens are mapped to \texttt{<unk>}. After
        capping, the actual learnt vocabularies contain
        $7{,}835$ German source tokens and $5{,}895$ English target
        tokens. Sequences are padded to a maximum of 50 tokens, and
        decoder targets are wrapped in \texttt{<sos>}/\texttt{<eos>}
        markers.
  \item \textbf{MarianMT pipeline:} the pretrained
        \texttt{Helsinki-NLP/opus-mt-de-en} tokenizer
        (SentencePiece-based) is applied directly, requiring no
        additional preprocessing.
\end{itemize}

\subsection{Models}

\subsubsection{Model A — Seq2Seq with Bahdanau Attention (Trained from Scratch)}

A bidirectional LSTM encoder maps source tokens to a context tensor; the
decoder is a unidirectional LSTM with Bahdanau attention
\cite{bahdanau2015neural}, where the attention layer computes a soft
alignment $\alpha_t$ over encoder states at every decoding step:

\[
\alpha_t = \mathrm{softmax}\bigl(v^{\top}\tanh(W[h_t^{\text{dec}};h_{1:T}^{\text{enc}}])\bigr),
\quad
c_t = \sum_{i=1}^{T}\alpha_{t,i}\, h_i^{\text{enc}}.
\]

The output projection conditions on the LSTM state, the attention
context, and the decoder embedding, following the standard formulation.
Hyperparameters are summarised in Table~\ref{tab:s2s_hp}; the trained
model has $12.12$ million parameters.

\begin{table}[h]
\centering
\caption{Seq2Seq+Attention Hyperparameters}
\label{tab:s2s_hp}
\begin{tabular}{ll}
\toprule
Hyperparameter      & Value \\
\midrule
Embedding dimension & 256 \\
Hidden dimension    & 256 (encoder bidirectional → 512 effective) \\
Dropout             & 0.30 \\
Batch size          & 32 \\
Epochs              & 8 \\
Optimiser / LR      & Adam / $10^{-3}$, StepLR (step 3, $\gamma=0.5$) \\
Teacher-forcing     & 0.50 \\
Gradient clipping   & 1.0 \\
Decoding            & Greedy \\
Total parameters    & 12.12 M \\
\bottomrule
\end{tabular}
\end{table}

\subsubsection{Model B — MarianMT (Pretrained Transformer)}

The \texttt{Helsinki-NLP/opus-mt-de-en} checkpoint
\cite{tiedemann2020opusmt} is a 6-layer encoder–decoder Transformer
trained on the OPUS corpora across many language pairs. The model is
used out-of-the-box for inference (no additional fine-tuning); generation
uses beam search with $B=4$ and a maximum output length of 128
sub-tokens.

\subsection{Evaluation Metrics}

We evaluate translations against the gold English captions using a
four-metric suite chosen to capture complementary aspects of quality:

\begin{itemize}
  \item \textbf{BLEU} \cite{papineni2002bleu} via \texttt{sacrebleu}.
        Standard $n$-gram precision with brevity penalty; reproducible
        exactly across toolchains.
  \item \textbf{METEOR} \cite{banerjee2005meteor}. Aligns hypotheses to
        references using stemming and WordNet synonyms.
  \item \textbf{ChrF} \cite{popovic2015chrf} via \texttt{sacrebleu}.
        Character $n$-gram F-score; particularly informative for
        morphologically rich languages because it credits stem matches
        even when affixes differ.
  \item \textbf{BERTScore F1} \cite{zhang2020bertscore} computed with a
        \texttt{distilbert-base-uncased} backbone (for memory
        efficiency). Captures semantic equivalence in contextual
        embedding space.
\end{itemize}

\subsection{Results}

\begin{table}[h]
\centering
\caption{Multi30k Test-Set Translation Results (de→en, n=1000)}
\label{tab:q4_results}
\begin{tabular}{lcc}
\toprule
Metric            & Seq2Seq+Attention & MarianMT \\
\midrule
BLEU              & 27.69 & \textbf{39.66} \\
ChrF              & 49.34 & \textbf{60.60} \\
METEOR            & 0.6267 & \textbf{0.7394} \\
BERTScore F1      & 0.8882 & \textbf{0.9333} \\
\bottomrule
\end{tabular}
\end{table}

MarianMT outperforms the from-scratch Seq2Seq across every metric. The
margin is particularly pronounced on BLEU ($+11.97$ points) and ChrF
($+11.26$ points), which reward exact $n$-gram matches and are
therefore especially sensitive to fluency and lexical accuracy. The
narrower margin on BERTScore (still $+0.045$) indicates that Seq2Seq
captures most of the semantic content but expresses it less fluently.

\subsection{Qualitative Examples}

\begin{table}[h]
\centering
\caption{Sample Outputs (test indices 0, 100, 500)}
\label{tab:q4_qual}
\small
\begin{tabular}{p{0.16\textwidth}p{0.74\textwidth}}
\toprule
\multicolumn{2}{l}{\textbf{Example 1 (test index 0)}} \\
\midrule
Source (de)  & Ein Mann mit einem orangefarbenen Hut, der etwas anstarrt. \\
Reference    & A man in an orange hat starring at something. \\
Seq2Seq+Attn & a man in an orange hat is something something . \\
MarianMT     & A man with an orange hat staring at something. \\
\midrule
\multicolumn{2}{l}{\textbf{Example 2 (test index 100)}} \\
\midrule
Source (de)  & Eine glückliche Frau bereitet in einem Coffee-Shop eine Erfrischung zu. \\
Reference    & A happy woman is preparing a refreshment at a coffee shop. \\
Seq2Seq+Attn & a happy is preparing a a a a a a a a . \\
MarianMT     & A happy woman prepares a refreshment in a coffee shop. \\
\midrule
\multicolumn{2}{l}{\textbf{Example 3 (test index 500)}} \\
\midrule
Source (de)  & Ein Tisch voller gerahmter Bilder, ein Freiluftmarkt. \\
Reference    & A table full of pictures in frames, at an outdoor market. \\
Seq2Seq+Attn & a table full of pictures pictures . \\
MarianMT     & A table full of framed pictures, an outdoor market. \\
\bottomrule
\end{tabular}
\end{table}

Across these three samples, three patterns are consistently observed.

\paragraph{Repetition collapse in Seq2Seq.}
Example 100 reveals a classic failure mode of greedy decoding in small
recurrent models: the model gets stuck on the article \textit{a} and
emits it eight times in succession. This is the well-documented
``repetition trap'' that motivated the development of coverage
mechanisms and beam search with length-penalty constraints. The same
phenomenon appears in Example 1 (\textit{``something something''}) and
Example 3 (\textit{``pictures pictures''}).

\paragraph{Vocabulary clipping artefacts.}
Example 100 also shows the Seq2Seq model dropping the noun
\textit{Frau} (woman): the German source word likely received a
sufficient embedding, but the target-side decoder failed to reproduce
\textit{woman} in this context. MarianMT's SentencePiece tokenizer and
larger pre-training corpus produce \textit{woman} fluently.

\paragraph{Surface fluency.}
Even when the Seq2Seq output is semantically close, it can produce
slightly ungrammatical English (article omission, lowercased proper
sentence starts). MarianMT outputs are uniformly fluent because the
decoder language model has been trained on orders of magnitude more
data.

\subsection{Discussion}

\paragraph{What each metric reflects.}
BLEU and ChrF reward surface-level overlap, with BLEU penalising
paraphrases harshly and ChrF being more forgiving thanks to
character-level matching --- particularly relevant for German
morphology. METEOR partially recovers paraphrases through stemming and
synonymy, and tends to track human judgement more closely than BLEU
alone. BERTScore is the most permissive: it credits semantic
equivalence even when surface forms differ substantially, which is why
the Seq2Seq model retains a relatively high BERTScore ($0.8882$)
despite its low BLEU ($27.69$).

\paragraph{Architectural trade-offs.}
The Seq2Seq+attention model has $12.12$M parameters, trains in roughly
ten minutes on M2 hardware, and runs inference in seconds. Its weakness
is data efficiency: with only 29k training pairs and randomly
initialised embeddings, it cannot match the quality of a model that has
seen orders of magnitude more bilingual data. MarianMT inherits roughly
$78$M parameters of pretrained knowledge, handles open-vocabulary
inputs through SentencePiece, and uses self-attention rather than
recurrence, all of which make it both more accurate and more robust.

The trade-off is in resource use and architectural transparency: a
hand-written Seq2Seq+attention model is a valuable pedagogical artefact
precisely because every component is visible and modifiable, whereas
the pretrained Marian checkpoint is a black box that one consumes
rather than understands.