% ─────────────────────────────────────────────────────────────────
% CENG 467 Midterm – Question 5: Language Modeling
% Student ID: 310201050
% ─────────────────────────────────────────────────────────────────

\section{Question 5: Language Modeling}

\subsection{Dataset}

The WikiText-2 corpus \cite{merity2017wikitext} is selected for this
investigation. WikiText-2 contains approximately 2 million tokens drawn
from a curated set of verified-good Wikipedia articles. The corpus is
preprocessed into fixed train ($2{,}051{,}910$ tokens), validation
($213{,}886$ tokens) and test ($241{,}211$ tokens) splits, and
words occurring fewer than three times in the training data have been
replaced with the \texttt{<unk>} symbol. These properties make
WikiText-2 the canonical small benchmark for language modelling
experiments and a natural choice for an Apple~M2 8\,GB development
setup.

The vocabulary is constructed directly from the training tokens
($33{,}277$ unique types) and shared by both models, ensuring that
perplexity numbers are directly comparable.

\subsection{Models}

\subsubsection{Model 1 — Trigram with Add-k Smoothing}

The first model is a classical trigram language model with add-$k$
smoothing. Letting $c(\cdot)$ denote training-corpus counts, the
probability of token $w_3$ given two-word context $(w_1, w_2)$ is

\[
P_{\text{add-}k}(w_3 \mid w_1, w_2) =
\frac{c(w_1, w_2, w_3) + k}{c(w_1, w_2) + k\,V},
\]

where $V$ is the vocabulary size and $k = 0.01$. This small $k$ assigns
non-zero probability to unseen trigrams while disturbing the
distribution over observed contexts only slightly. Compared with
classical add-1 (Laplace) smoothing it preserves more of the empirical
signal in high-count contexts.

The model has no learned parameters in the usual sense; estimation is a
single pass over the training corpus. Generation is performed by
sampling from the smoothed distribution at each step.

\subsubsection{Model 2 — 2-Layer LSTM Language Model}

The second model is a two-layer LSTM \cite{hochreiter1997lstm} trained
end-to-end with cross-entropy loss using truncated back-propagation
through time (BPTT). The trained model has $15.93$ million parameters.

\begin{table}[h]
\centering
\caption{LSTM Language Model Hyperparameters}
\label{tab:lstm_lm_hp}
\begin{tabular}{ll}
\toprule
Hyperparameter   & Value \\
\midrule
Embedding dimension      & 192 \\
Hidden dimension         & 256 \\
Number of LSTM layers    & 2 \\
Dropout                  & 0.45 \\
BPTT context length      & 35 tokens \\
Batch size               & 20 (BPTT layout) \\
Epochs                   & 6 \\
Optimiser / LR           & Adam / $10^{-3}$ \\
LR schedule              & StepLR (step 2, $\gamma=0.5$) \\
Gradient clipping        & 0.25 \\
Total parameters         & 15.93 M \\
\bottomrule
\end{tabular}
\end{table}

The hidden state is detached at every BPTT window so gradients do not
back-propagate across windows, which both stabilises training and keeps
memory consumption bounded. At inference, a fresh zero hidden state
followed by the prompt tokens primes the model before sampling begins.

\subsection{Evaluation: Perplexity}

Perplexity is the exponentiated per-token cross-entropy:

\[
\mathrm{PPL}(w_1,\dots,w_N) = \exp\Bigl(-\tfrac{1}{N}\sum_{i=1}^{N} \log P(w_i \mid w_{<i})\Bigr).
\]

It is the standard intrinsic measure of language-model quality; lower
values indicate that the model assigns higher probability mass to
held-out text.

\begin{table}[h]
\centering
\caption{WikiText-2 Perplexity (lower is better; seed 3102)}
\label{tab:q5_results}
\begin{tabular}{lcc}
\toprule
Model                            & Validation PPL & Test PPL \\
\midrule
Trigram (add-$k$, $k = 0.01$)    & 3868.40 & 3587.97 \\
2-layer LSTM                     & \textbf{389.69} & \textbf{353.41} \\
\bottomrule
\end{tabular}
\end{table}

The LSTM achieves perplexity approximately \emph{ten times lower} than
the trigram baseline ($353.41$ vs.\ $3587.97$ on test). This dramatic
gap quantifies the benefit of distributed representations and longer
effective context for next-token prediction.

\subsection{Text Generation Samples}

Both models are seeded with three identical short prompts and generate
25 follow-up tokens by sampling from their respective distributions
(temperature $\tau = 0.9$ for the LSTM; raw smoothed distribution for the
trigram).

\begin{table}[h]
\centering
\caption{Generated Continuations (excerpts)}
\label{tab:q5_gen}
\small
\begin{tabular}{p{0.13\textwidth}p{0.78\textwidth}}
\toprule
Prompt    & ``the city of'' \\
Trigram   & \textit{... Zagreb. However, German mycologist Rolf Singer listed three subspecies, although it was then tied with Tower of Terror, with Julio Jones} \\
LSTM      & \textit{... achieving to October the kings before the soul. There are full, was \texttt{<unk>}, almost serving as the Bruins as a short school} \\
\midrule
Prompt    & ``in the year'' \\
Trigram   & \textit{... . Twenty @-@ four @-@ issue story arc to be in a Pepsi television advertisement in the north of the young characters into one of} \\
LSTM      & \textit{... – two season and the 17th of arguing and fought on Scotland and ₹ 85 (1 @.@) mi; the storm – 16} \\
\midrule
Prompt    & ``she said that'' \\
Trigram   & \textit{... the problem in many cases from several teams that were arrested and killed four young black girls and football manager. Lawton and Sewell's} \\
LSTM      & \textit{... it is `` that related is while deal as his political titles '', by the first name on six side, being peaking by} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Trigram outputs.}
Locally fluent two- and three-word fragments appear because every chosen
word has been observed immediately following the preceding bigram.
However, coherence collapses beyond the trigram window: the topic
drifts unpredictably across very different domains within a single
sample (e.g.\ from \textit{German mycologist} to \textit{Tower of
Terror} to \textit{Julio Jones}), and named entities are introduced
without context. Topical coherence is bounded by the length of the
$n$-gram context and breaks immediately at that boundary.

\paragraph{LSTM outputs.}
The LSTM produces longer fluent spans with consistent grammatical
structure (\textit{``almost serving as the Bruins as a short school''}
preserves clausal structure even where content is incoherent). Failure
modes are different from the trigram model's: rather than topic drift,
the LSTM tends to combine plausible noun phrases in semantically
implausible ways, and occasionally produces tokens such as \texttt{₹},
\texttt{@-@}, and \texttt{<unk>} that are artefacts of the WikiText-2
preprocessing rather than genuine language. These behaviours are
characteristic of LM-decoded text trained with maximum-likelihood
objectives on a small corpus.

\subsection{Discussion}

\paragraph{Why the LSTM dominates on perplexity.}
The trigram model estimates one parameter per observed $(w_1, w_2,
w_3)$ triple, which suffers from severe data sparsity: even on
WikiText-2's $\sim$2M training tokens, the vast majority of
theoretically possible trigrams remain unobserved, so add-$k$ smoothing
must allocate substantial probability mass to all of them uniformly.
The LSTM, in contrast, learns a continuous distributed representation
in which similar contexts produce similar predictions, generalising
automatically across rare or unseen surface patterns. The
$10\times$ perplexity gap quantifies this advantage.

\paragraph{Effective context.}
Although the LSTM is in principle able to condition on arbitrarily long
histories, in practice the information that survives many recurrent
updates degrades. Empirical studies have shown effective context
lengths of a few hundred tokens for deep LSTMs; modern Transformers use
direct attention to circumvent this limitation, which is the main
reason they have largely supplanted recurrent architectures for
language modelling. The trigram model's effective context is fixed at
two tokens, which is why its perplexity remains in the four-digit range
even after smoothing.

\paragraph{Trade-offs.}
The trigram model trains in seconds, requires no GPU, and is perfectly
transparent --- every probability is a counted fraction. The LSTM
requires multiple training epochs and approximately $16$M parameters,
but produces dramatically lower perplexity and generates more coherent
samples. For applications that need quick statistical baselines (corpus
statistics, code-switch detection, forced alignment) a smoothed
$n$-gram remains a perfectly serviceable choice; for any application
that requires generation quality, the neural model is preferable.